---
title: "Master script for postfire analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Source functions, get data and plot

First we'll _source()_ (i.e. "run all code in") the scripts with the functions we made. Then we'll set the URL, read in the data with _download.NDVI()_, and plot it with _plot.NDVI()_.

```{r}
## Load required functions by running source() on the individual function files
if(file.exists("01_download.NDVI.R")) source("01_download.NDVI.R")
if(file.exists("02_plot.NDVI.R"))     source("02_plot.NDVI.R")
if(file.exists("03_negexp.R"))        source("03_negexp.R")
## Download NDVI data
URL = "https://raw.githubusercontent.com/jslingsby/BIO3019S_Ecoforecasting/master/data/modisdata.csv"
dat <- download.NDVI(URL)
# Convert "calendar_date" to postfire age in days since fire - assuming the first date in the times eries is the time of the fire 
dat$age <- (as.numeric(dat$calendar_date) - min(as.numeric(dat$calendar_date), na.rm = T))/365.25
## Plot overall NDVI time series
plot.NDVI(dat)
```

<br>

Q1: This plot suggests that Fynbos greenness (NDVI) as observed from satellite saturates with time since fire. Why do you think it saturates rather than increasing linearly with time?

>*Answer 1:* There are likely seasonal changes in NDVI which cause the saturation observed in the plot. As NDVI is a level of greeness, for many plant species, NDVI might decrease in winter months (when leaves drop or are blocked from satellite view from factors such as snow) and peak in summer. Many Fynbos plants do not drop their leaves in winter so there is likely other reasons for the observed saturation evident on the plot. Apart from physiological changes of Fynbos leaf that occur during a year which could cause NDVI saturation, atmospheric conditions could also come into play in affecting satellite readings. MODIS is a passive sensor, meaning that although the greeness of certain species may not change seasonally,  if there is mist or haze, the MODIS satellite reading of NDVI would not be accurate as the sensors can only detect when there is naturally occurring energy. 
<br>

### 2. Fit models using Non-linear Least Squares (NLS)

Now we'll fit the simple and full negative exponential models using Non-linear Least Squares (NLS).

First the simpler model:

```{r}
## Simple model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexp <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda)),
                  data = dat, start = par, trace = F, 
                  control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexp)
```

<br>

And let's look at the model summary with parameter estimates

```{r}
# print model summary
summary(fit_negexp)
```

<br>

Now the full model:

```{r}
## Full model
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpS <- nls(NDVI ~ alpha + gamma * (1 - exp(- age/lambda))
                   + A*sin(2*pi*age + (phi + pi/6*(3 - 1))), 
                   data = dat, start = par, trace = F, 
                   control = nls.control(maxiter = 500))
# plot
plot.NDVI(dat = dat, fit = fit_negexpS)
```


```{r}
# print model summary
summary(fit_negexpS)
```

<br>

Lots more parameters...

Q2: How do the estimates for the common parameters compare?

>*Answer 2:* In both models all included parameters have a significant effect on the response variable (Pr <0.001). In terms of parameter estimates, the shared parameters between both models are alpha, gamma and lambda. The estimate values for alpha and lambda both decrease from the simple to the full model (from 0.25107 to 0.207522 and from 1.17687 to 0.989154, respectively). This may be due to other parameters included in the full model accounting for some of the variation accredited to these parameters in the simple model. The estimate value of the gamma parameter increase from 0.32371 to 0.364746 from the simple to full model. 

<br>

### 3. Compare NLS models using ANOVA

Modelers often want to know which of a set of models are better. One way to do this when comparing nested* models using least squares is using analysis of variance (ANOVA). In this case the `anova()` function will take the model objects as arguments, and return an ANOVA testing whether the full model results in a significant reduction in the residual sum of squares (and thus is better at capturing the data), returning an F-statistic, Degrees of Freedom (the difference in the number of parameters between the models) and p-value.

*i.e. one model is a subset of the other, as in our case

```{r}
anova(fit_negexp, fit_negexpS)
```

<br>

Q3: Which model is better?

>*Answer 3:* In using the output from the anova() function, one can tell that Model 2 (i.e. the full NLS) provide a significantly better fit to the data compared to Model 1 (simple NLS), as Pr = 4.12e-14. 

<br>

Q4: How many degrees of freedom are there in this ANOVA and why (i.e. what are they)?

>*Answer 4:* There are 2 degrees of freedom and they are the A and phi parameters (i.e. the parameters in Model 2 which are not in Model 1).

<br>

### 4. Fit models using Maximum Likelihood Estimation (MLE)

First let's fit the simpler model:

```{r}
## Fit the simpler model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5)
# fit model
fit_negexpMLE <- fit.negexp.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexp(fit_negexpMLE$par,dat$age), col = 'skyblue', lwd = 3)
```


```{r}
fit_negexpMLE
```

<br>

Then the full model:

```{r}
## Fit the full model using MLE
# set parameters
par <- c(alpha = 0.2, gamma = 0.4, lambda = 0.5, A = 0.6, phi = 0)
# fit model
fit_negexpMLES <- fit.negexpS.MLE(dat, par)
# plot
plot.NDVI(dat)
# add curve with MLE parameters
lines(dat$age, pred.negexpS(fit_negexpMLES$par,dat$age), col = 'skyblue', lwd = 3)
```

```{r}
fit_negexpMLES
```

<br>

### 5. Compare MLE models using Akaike's information criterion (AIC)

Note that we can't compare our MLE models using ANOVA because our custom functions do not return full model fits like the `nls()` function - only the parameter estimates, negative log-likelihoods and a few other diagnostics.

Another way to compare models (and probably the most common) is using the Akaike information criterion (AIC), which is an estimator of prediction error (i.e. relative quality) of statistical models for a given set of data. 

The formula for the Akaike information criterion is:

$AIC = 2K -2(ln(L))$

Where:

- $k$ = the number of estimated parameters in the model
- $L$ = maximum value of the likelihood function for the model

Since we have our negative log likelihoods (i.e. $-ln(L)$ in the formula above), we can calculate the AICs and compare them.

```{r}
AIC_simple = 6 + 2*fit_negexpMLE$value
AIC_simple
AIC_full = 6 + 2*fit_negexpMLES$value
AIC_full
```

<br>

When comparing models, the lower the AIC the better, and in general a difference in AIC of 3 or more is analagous to the models being significantly different at an $\alpha$ of $p < 0.05$.

```{r}
AIC_simple - AIC_full
```

<br>

Q5: Is there a preferred model and if so, which one?

>*Answer 5:* Yes. The full model is preferred as it has a lower AIC value compared to the simple model (AIC_simple - AIC_full = 266957.8; this indicates that the models are significantly different).

<br>

The nice thing about AIC is that the models you compare do not have to be nested like they do for ANOVA, as long as the data are the same. There are a few other constraints however... 

Here are the AIC scores for our pair of NLS models:

```{r}
AIC(fit_negexp, fit_negexpS)
```

<br>

You'll notice that these are completely different to the AICs for the MLE models...

Q6: Why is it not okay to compare the AIC of these NLS models with the AIC of the MLE models? Hint: type `?AIC` into the R console and do some reading.

>*Answer 6:* Theoretically, AIC requires the maximisation of log-likelihood, however AIC can be calculated for models that are not fitted by maximum likelihood. In this case, AIC is calculated for MLE  on the basis of maximised likelihood and NLS is not. Thus, their AIC values should not be compared.

<br>
